---
title: "The Effect of Gross Square Footage on the Sale Prices of Single Family Homes in New York City"
author: "Aleksei Ostlund"
date: "03/10/2021"
output: 
  html_document: 
    theme: cerulean
    code_folding: hide
    highlight: haddock
---
# Introduction

The data sets used in this project are made publicly available by the City of New York. The data contained represents property sales for a 12-month rolling time period. For the purposes of this analysis, the time frame was September 2020 - August 2021.  

The purpose of this project is to measure how well the size of housing explains the sale price across New York City and its boroughs. The housing type chosen for this analysis is A5: attached or semi-detached single family homes.

The question of how well gross square footage predicts the sale price in each borough will be addressed by using linear regression modeling.  



__Load packages__

```{r  message=FALSE}
library(magrittr) 
library(stringr) 
library(dplyr) 
library(readr) 
library(ggplot2)
library(hrbrthemes)
library(tidyverse)
```

__Data import__

```{r remedy001}
library(readxl)
manhattan_data <- read_xlsx("rollingsales_manhattan.xlsx", skip = 4)
bronx_data <- read_xlsx("rollingsales_bronx.xlsx", skip = 4)
brooklyn_data <- read_xlsx("rollingsales_brooklyn.xlsx", skip = 4)
queens_data <- read_xlsx("rollingsales_queens.xlsx", skip = 4)
si_data <- read_xlsx("rollingsales_statenisland.xlsx", skip = 4)

```

__Combining rows from all 5 tables__

```{r remedy002}

nyc_property_data <- bind_rows(bronx_data, brooklyn_data, manhattan_data, queens_data, si_data)
# drop other tables 
remove(bronx_data, brooklyn_data, manhattan_data, queens_data, si_data)

# Change borough names from numbers to words
nyc_property_data <- nyc_property_data %>%
  mutate(BOROUGH=case_when(BOROUGH == 1 ~ 'MANHATTAN', 
                           BOROUGH == 2 ~ 'BRONX',
                           BOROUGH == 3 ~ 'BROOKLYN',
                           BOROUGH == 4 ~ 'QUEENS',
                           BOROUGH == 5 ~ 'STATEN_ISLAND',
                           TRUE ~ 'na'))

```

__Convert columns to lower case, remove spaces, duplicates, unneeded columns__

```{r remedy003}

colnames(nyc_property_data) %<>% str_replace_all("\\s", "_") %>% tolower()

# Convert capitalized fields to title case
nyc_property_data <- nyc_property_data %>%
  mutate(borough = str_to_title(nyc_property_data$borough, locale = 'en')) %>%
  mutate(neighborhood = str_to_title(nyc_property_data$neighborhood, locale = 'en')) %>%
  mutate(building_class_category = str_to_title(nyc_property_data$building_class_category, locale = 'en')) %>%
  mutate(address = str_to_title(nyc_property_data$address, locale = 'en'))

# Remove possible duplicates
nyc_property_data <- nyc_property_data %>%
  distinct()
# Drop irrelevant columns
nyc_property_data <- subset(nyc_property_data, select = -easement )

```

__Drop sales within families (assumed as under 10k), <150 sqft entries (The NYC Building Code requires all dwellings to be >150), and NA entries for sales/gross square feet columns__

```{r remedy004}

nyc_property_data <- nyc_property_data %>%
  filter(sale_price>10000) %>%
  filter(gross_square_feet>=150) %>%
  drop_na(sale_price, gross_square_feet) %>%
  arrange(borough, neighborhood)

# Export to CSV
write_csv(nyc_property_data, file='nycsales_cleaned.csv')

```

__Filter for one family attached or semi-detached housing__

```{r remedy005}

one_family_attached <- nyc_property_data %>%
  filter(building_class_at_time_of_sale == 'A5')

```

__Scatterplots__

```{r warning=FALSE, message=FALSE, fig.height=7, fig.width=10}


options(scipen = 100)

ggplot(data= one_family_attached, mapping = aes(x= sale_price, y= gross_square_feet, color=borough))+
  geom_point()+
  xlim(0,7500000)+
  geom_smooth(method = 'lm', se= FALSE, aes(group = 1)) + theme(axis.title = element_text(size = 12),
    plot.title = element_text(face = "bold")) +labs(title = "The effect of gross square footage on sale price for attached single family homes in NYC",
    x = "Sale price ($)", y = "Gross Square Footage",
    colour = "Borough") +
  theme_minimal()+
  ylim(0,5000) + theme(plot.title = element_text(face = "bold")) +labs(title = "Sale Price vs Gross Square Footage in NYC")+labs(subtitle = "Single family homes, attached or semi-detached")
```

```{r warning=FALSE, message=FALSE, fig.width=10,fig.height=11}
# Scatterplot by borough
options(scipen = 100)
library(hrbrthemes)
ggplot(data= one_family_attached, mapping = aes(x= sale_price, y= gross_square_feet))+
  geom_point(color = 'red')+
  facet_wrap(~borough, ncol = 1, scales = 'free')+
  scale_y_continuous()+
  geom_smooth(method = 'lm', se= FALSE, color = 'blue') +
  theme_minimal() + theme(plot.title = element_text(face = "bold")) +labs(title = "Sale Price vs Gross Square Footage by Borough",
    x = "Sale Price", y = "Gross Square Footage")+labs(subtitle = "Single family homes, attached or semi-detached") + theme(legend.position = "none")+labs(x = "Sale Price ($)")

```

Top sale in Manhattan is the Herbert M Strouse House, likely the largest private residence in Manhattan. It has a celebrity factor.
I did not consider this an outlier as it is still a legitimate data point.

__Generate linear regression models__

```{r remedy007}

one_family_lm <- lm(sale_price ~ gross_square_feet,data = one_family_attached)
summary(one_family_lm)

```

The model has a sufficiently low p value (below 0.001) and high t value so that we can say the two variables are related.
The r squared score for the entirety of New York is a moderate 0.4422. 

__Generate separate linear regression models for each borough.__

```{r}

library(broom)
library(tidyr)
library(purrr)
borough_lms <- one_family_attached %>%
  group_by(borough) %>%
  nest() %>%
  mutate(linear_model = map(.x = data,
                            .f = ~lm(sale_price ~ gross_square_feet,
                            data = .)))

```

__Generate tidy coefficients and regression summaries__

```{r}

borough_lms <- borough_lms %>%
  mutate(tidy_coefficients = map(.x = linear_model,
                                 .f = tidy,
                                 conf.int = TRUE))

#Unnest
tidy_unnested <- borough_lms %>%
  select(borough, tidy_coefficients) %>%
  unnest(cols = tidy_coefficients)

tidy_unnested %>%
  filter(term == 'gross_square_feet') %>%
  rename(slope = estimate) %>%
  select(borough, term, slope) %>%
  arrange((desc(slope))) %>%
  print()

```

When looking at the slopes, we can see that Manhattan has the biggest price
increase per sq foot followed by Brooklyn. Other boroughs are noticeable behind.


__Generate regression summaries__

```{r remedy008}

borough_regressions <- one_family_attached %>%
  group_by(borough) %>%
  nest() %>%
  mutate(linear_regression = map(.x = data,
                                 .f = ~lm(sale_price ~ gross_square_feet, data =.
                                       ))) %>%
  mutate(tidy_regressions = map(.x = linear_regression,
                                .f = glance,
                                conf.int = TRUE)) %>%
  select(borough, tidy_regressions)%>%
  unnest(cols = tidy_regressions)

borough_regressions %>% arrange(desc(adj.r.squared)) %>% 
  select(borough, adj.r.squared) %>% 
  print()

```

__Neighborhoods with best fit for linear model__

```{r remedy009}


# Remove neighborhoods with less than 10 sales
neighborhoods_count <- one_family_attached %>%
  count(neighborhood) %>%
  filter(n>9)
# New dataset
neighborhoods_cleaned <- one_family_attached %>%
  filter(neighborhood %in% neighborhoods_count$neighborhood)

# Linear regressions by neighborhood
neighborhood_regressions <- neighborhoods_cleaned %>%
  group_by(borough, neighborhood) %>%
  nest() %>%
  mutate(linear_model = map(.x = data,
             .f = ~lm(sale_price ~ gross_square_feet, data = .))) %>%
  mutate(tidy_regressions = map(.x = linear_model,
                                 .f = glance,
                                 conf.int = TRUE)) %>%
  select(borough, neighborhood, tidy_regressions) %>%
  unnest(cols=tidy_regressions)

# Remove high p values and filter for adj r squared
significant_neighborhood_regressions <- neighborhood_regressions %>%
  filter(p.value <0.05) %>%
  filter(adj.r.squared > 0.5) %>%
  arrange(desc(adj.r.squared))

print(significant_neighborhood_regressions)

```

__Graph of neighborhoods where gross square footage is the best predictor of sale price__

```{r warning=FALSE, message=FALSE, fig.width= 10}

# Graph these neighborhoods

neighborhood_graphs <- one_family_attached %>%
  filter(neighborhood == 'Ocean Parkway-South' |
           neighborhood == 'Gravesend' |
           neighborhood == 'Windsor Terrace' |
           neighborhood == 'Mill Basin' |
           neighborhood == 'Sunset Park' |
           neighborhood == 'Far Rockaway')

ggplot(data=neighborhood_graphs, mapping= aes(x=sale_price, y=gross_square_feet))+
  geom_point(color='green') +
  theme_minimal() +
  geom_smooth(method = 'lm', se=FALSE) +
  facet_wrap(~neighborhood, ncol = 2) + theme(axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold")) +labs(title = "Sale Price vs Gross Square Feet for Select Neighborhoods",
    x = "Sale Price ($)", y = "Gross Square Feet",
    subtitle = "Attached or semi-detached single family homes")

```

# Conclusion
New York City as a whole had a moderate-weak correlation between gross square feet and sale price. Looking at the r-squared values for all 5 boroughs individually it appears that Manhattan has a strong value and Brooklyn has a low-moderate one. The other 3 have low values. Furthermore, Manhattan had the largest sale price increase per square foot. Brooklyn was second while the other 3 boroughs did not have such large increases.

These values indicate that in the Bronx, Queens, and Staten Island, other factors have a larger impact on pricing of attached/semi-detached single family homes. These could be location, proximity to transport, or the quality of the interior. Finally, the values for Manhattan needs to be investigated more as there are only 10 sales of this property type in the past year. Data from previous years could be used to compare r squared values.

When picking out specific neighborhoods, there were 6 that had an r squared value greater than 0.5 while having suitable p and t values. When looking at sales for the chosen housing class in these select neighborhoods gross square footage can be seen as a fairly reliable predictor or the sale price.
